---
title: "Limits, colimits, and graph rewriting in catlab"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Limits, colimits, and graph rewriting in catlab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `catlab` package provides categorical algebra tools for R, built on
the `acsets` package for attributed C-sets. This vignette demonstrates:

1. **Limits** — products, pullbacks, equalizers, terminal objects
2. **Colimits** — coproducts, pushouts, coequalizers, initial objects
3. **Homomorphism finding** — pattern matching in graphs
4. **DPO graph rewriting** — algebraic graph transformation via double pushout
5. **Data migration** — delta (pullback) and sigma (pushforward) migration

```{r setup}
library(catlab)
library(acsets)
```

## Limits and colimits

Limits and colimits are universal constructions in category theory. For
ACSets (attributed C-sets), these operations work pointwise on each
object type.

### Terminal and initial objects

The **terminal object** has exactly one part per object type. Every ACSet
has a unique morphism to it. The **initial object** has zero parts.

```{r terminal-initial}
t <- terminal(SchGraph)
cat("Terminal graph: V =", nv(t), ", E =", ne(t), "\n")

i <- initial(SchGraph)
cat("Initial graph: V =", nv(i), ", E =", ne(i), "\n")

# Unique morphism to terminal
g <- path_graph(3)
alpha <- to_terminal(g)
cat("Morphism maps all", nv(g), "vertices to vertex 1\n")
cat("Natural?", is_natural(alpha), "\n")
```

### Coproducts (disjoint union)

The coproduct of two ACSets is their disjoint union, with injection
morphisms tracking where each part came from.

```{r coproduct}
g1 <- path_graph(2)   # 1 → 2
g2 <- cycle_graph(3)  # 1 → 2 → 3 → 1

cp <- coproduct(g1, g2)
cat("g1: V =", nv(g1), ", E =", ne(g1), "\n")
cat("g2: V =", nv(g2), ", E =", ne(g2), "\n")
cat("Coproduct: V =", nv(cp$coproduct), ", E =", ne(cp$coproduct), "\n")
cat("Injection 1 (V):", cp$inj1@components$V, "\n")
cat("Injection 2 (V):", cp$inj2@components$V, "\n")
```

### Products (categorical product)

The product of two graphs contains all pairs of vertices. An edge
exists in the product only when both component graphs have edges
between the corresponding vertices.

```{r product}
g1 <- path_graph(2)  # 1 → 2
g2 <- path_graph(2)  # 1 → 2

p <- product(g1, g2)
cat("Product of two edges: V =", nv(p$product), ", E =", ne(p$product), "\n")
cat("Projection 1 (V):", p$proj1@components$V, "\n")
cat("Projection 2 (V):", p$proj2@components$V, "\n")
```

The product has 4 vertices (pairs) but only 1 edge: the pair of edges
$(1 \to 2, 1 \to 2)$ maps to vertex pair $(1,1) \to (2,2)$.

### Pushouts (gluing)

The pushout glues two ACSets along a common substructure. Given morphisms
$f: A \to B$ and $g: A \to C$, the pushout identifies $f(a)$ with $g(a)$
for each part $a$ of $A$.

```{r pushout}
# Glue two edges at a shared vertex
A <- Graph(); add_vertices(A, 1)  # Shared vertex
B <- path_graph(2)                 # 1 → 2
C <- path_graph(2)                 # 1 → 2

# Share the source vertex
f <- ACSetTransformation(list(V = 1L, E = integer(0)), A, B)
g <- ACSetTransformation(list(V = 1L, E = integer(0)), A, C)

po <- pushout(f, g)
cat("Pushout: V =", nv(po$pushout), ", E =", ne(po$pushout), "\n")
# 2 + 2 - 1 = 3 vertices, 2 edges (the shared vertex is identified)
```

### Pullbacks (fiber product)

The pullback of a cospan $B \to D \leftarrow C$ consists of pairs
$(b, c)$ where $f(b) = g(c)$ — the "fiber product."

```{r pullback}
# D: 2 vertices; B: 3 vertices mapping {1,2→1, 3→2}; C: 2 vertices mapping {1→1, 2→2}
D <- Graph(); add_vertices(D, 2)
B <- Graph(); add_vertices(B, 3)
C <- Graph(); add_vertices(C, 2)

f <- ACSetTransformation(list(V = c(1L, 1L, 2L), E = integer(0)), B, D)
g <- ACSetTransformation(list(V = c(1L, 2L), E = integer(0)), C, D)

pb <- pullback(f, g)
cat("Pullback: V =", nv(pb$pullback), "\n")
cat("Pairs (B, C):")
for (i in seq_len(nv(pb$pullback))) {
  cat(sprintf(" (%d,%d)", pb$proj1@components$V[i], pb$proj2@components$V[i]))
}
cat("\n")
# Fiber over D=1: B parts {1,2} × C parts {1} = 2 pairs
# Fiber over D=2: B parts {3} × C parts {2} = 1 pair
```

### Equalizers and coequalizers

The **equalizer** of two parallel morphisms $f, g: A \to B$ is the
subobject of $A$ where $f$ and $g$ agree.

```{r equalizer}
A <- Graph(); add_vertices(A, 3)
B <- Graph(); add_vertices(B, 3)

# f: 1→1, 2→2, 3→3 (identity); g: 1→1, 2→1, 3→3 (merges 2 into 1)
f <- ACSetTransformation(list(V = c(1L, 2L, 3L), E = integer(0)), A, B)
g <- ACSetTransformation(list(V = c(1L, 1L, 3L), E = integer(0)), A, B)

eq <- equalizer(f, g)
cat("Equalizer: V =", nv(eq$equalizer), "\n")
cat("Included vertices:", eq$incl@components$V, "\n")
# Vertices 1 and 3 agree; vertex 2 doesn't
```

The **coequalizer** identifies elements in the codomain:

```{r coequalizer}
# Identify vertices 1 and 2 in B
A <- Graph(); add_vertices(A, 1)
B <- Graph(); add_vertices(B, 3)

f <- ACSetTransformation(list(V = 1L, E = integer(0)), A, B)
g <- ACSetTransformation(list(V = 2L, E = integer(0)), A, B)

coeq <- coequalizer(f, g)
cat("Coequalizer: V =", nv(coeq$coequalizer), "(from 3)\n")
cat("Quotient map:", coeq$proj@components$V, "\n")
```

## Homomorphism finding

A **homomorphism** between two ACSets is an ACSet transformation — a
structure-preserving map. The `find_homomorphism()` function uses
backtracking search to find morphisms from a pattern to a target.

```{r homomorphism-basic}
# Find an edge in a triangle
edge <- path_graph(2)
triangle <- cycle_graph(3)

h <- find_homomorphism(edge, triangle)
cat("Edge maps to triangle:\n")
cat("  V:", h@components$V, "\n")
cat("  E:", h@components$E, "\n")
cat("  Natural?", is_natural(h), "\n")
```

### Finding all homomorphisms

```{r homomorphism-all}
all_h <- find_all_homomorphisms(edge, triangle, monic = TRUE)
cat("Monic homomorphisms (edge → triangle):", length(all_h), "\n")
for (i in seq_along(all_h)) {
  cat(sprintf("  Match %d: V=[%s], E=[%s]\n", i,
    paste(all_h[[i]]@components$V, collapse = ","),
    paste(all_h[[i]]@components$E, collapse = ",")))
}
```

### Checking embeddability

```{r embeddability}
cat("Edge embeds in triangle?", is_homomorphic(edge, triangle, monic = TRUE), "\n")
cat("Triangle embeds in edge?", is_homomorphic(triangle, edge, monic = TRUE), "\n")
```

## Double pushout (DPO) graph rewriting

DPO rewriting is the categorical foundation of algebraic graph
transformation. A **rewrite rule** is a span $L \leftarrow I \rightarrow R$:

- **L** (left-hand side): the pattern to match
- **I** (interface): the structure preserved during rewriting
- **R** (right-hand side): the replacement

The DPO construction proceeds in two steps:

1. **Pushout complement**: Remove matched pattern, keeping interface → context K
2. **Pushout**: Glue replacement R to context K via interface → result H

```
     l         r
  L ← I → R
m ↓   ↓     ↓
  G ← K → H
```

### Example 1: Add an edge

Create a rule that adds an edge between two matched vertices.

```{r dpo-add-edge}
# Interface: 2 isolated vertices
I <- Graph(); add_vertices(I, 2)
# Pattern: 2 isolated vertices (same as interface)
L <- Graph(); add_vertices(L, 2)
# Replacement: edge 1 → 2
R <- path_graph(2)

l <- ACSetTransformation(list(V = c(1L, 2L), E = integer(0)), I, L)
r <- ACSetTransformation(list(V = c(1L, 2L), E = integer(0)), I, R)
add_edge_rule <- rule(l, r)

# Apply to a graph with 3 isolated vertices
G <- Graph(); add_vertices(G, 3)
H <- rewrite(add_edge_rule, G)
cat("Before: V =", nv(G), ", E =", ne(G), "\n")
cat("After:  V =", nv(H), ", E =", ne(H), "\n")
```

### Example 2: Delete an edge

```{r dpo-delete-edge}
# Interface: 2 isolated vertices
I <- Graph(); add_vertices(I, 2)
# Pattern: edge 1 → 2
L <- path_graph(2)
# Replacement: 2 isolated vertices
R <- Graph(); add_vertices(R, 2)

l <- ACSetTransformation(list(V = c(1L, 2L), E = integer(0)), I, L)
r <- ACSetTransformation(list(V = c(1L, 2L), E = integer(0)), I, R)
del_edge_rule <- rule(l, r)

# Apply to path graph 1 → 2 → 3
G <- path_graph(3)
H <- rewrite(del_edge_rule, G)
cat("Before: V =", nv(G), ", E =", ne(G), "\n")
cat("After:  V =", nv(H), ", E =", ne(H), "\n")
```

### Example 3: Add a vertex

```{r dpo-add-vertex}
I <- Graph()  # Empty interface
L <- Graph()  # Empty pattern (matches anywhere)
R <- Graph(); add_vertices(R, 1)  # Add a vertex

l <- ACSetTransformation(list(V = integer(0), E = integer(0)), I, L)
r <- ACSetTransformation(list(V = integer(0), E = integer(0)), I, R)
add_vertex_rule <- rule(l, r, monic = FALSE)

G <- path_graph(2)
H <- rewrite(add_vertex_rule, G)
cat("Before: V =", nv(G), ", E =", ne(G), "\n")
cat("After:  V =", nv(H), ", E =", ne(H), "\n")
```

### Finding all matches

Before rewriting, we can enumerate all valid matches:

```{r matches}
matches <- get_matches(del_edge_rule, cycle_graph(4))
cat("Delete-edge rule has", length(matches), "matches in a 4-cycle\n")
for (i in seq_along(matches)) {
  cat(sprintf("  Match %d: V=[%s], E=[%s]\n", i,
    paste(matches[[i]]@components$V, collapse = ","),
    paste(matches[[i]]@components$E, collapse = ",")))
}
```

### Gluing condition violations

DPO rewriting requires **gluing conditions** to be satisfied. Deleting
a vertex that has incident edges (not in the interface) violates the
dangling condition:

```{r gluing-violation}
# Try to delete a vertex from a graph where it has edges
I <- Graph()
L <- Graph(); add_vertices(L, 1)
R <- Graph()

l <- ACSetTransformation(list(V = integer(0), E = integer(0)), I, L)
r <- ACSetTransformation(list(V = integer(0), E = integer(0)), I, R)
del_vertex_rule <- rule(l, r, monic = FALSE)

G <- path_graph(2)  # 1 → 2
# Matching vertex 1 (which has an outgoing edge) should fail
tryCatch(
  rewrite_match(del_vertex_rule,
    ACSetTransformation(list(V = 1L, E = integer(0)), L, G)),
  error = function(e) cat("Expected error:", conditionMessage(e), "\n")
)
```

## Data migration

Data migration transforms ACSets between different schemas using functors.

### Delta migration (pullback)

Given a functor $F: C \to D$ and a $D$-set, delta migration produces
a $C$-set by pulling back along $F$.

```{r delta-migration}
# Reverse a graph: swap src and tgt
cat_G <- FinCat(schema = SchGraph)
reverse_F <- FinFunctor(
  ob_map = list(V = "V", E = "E"),
  hom_map = list(src = "tgt", tgt = "src"),  # Swap!
  dom = cat_G, codom = cat_G
)

g <- path_graph(3)  # 1 → 2 → 3
GraphType <- acset_type(SchGraph)
g_rev <- delta_migrate(reverse_F, g, GraphType)

cat("Original: src =", subpart(g, NULL, "src"), ", tgt =", subpart(g, NULL, "tgt"), "\n")
cat("Reversed: src =", subpart(g_rev, NULL, "src"), ", tgt =", subpart(g_rev, NULL, "tgt"), "\n")
```

### Sigma migration (left Kan extension)

Sigma migration is the left adjoint to delta — it pushes data forward
along a functor via colimits.

```{r sigma-migration}
# Combine two edge types into one graph
dom_sch <- BasicSchema(
  obs = c("V", "E1", "E2"),
  homs = list(
    hom("src1", "E1", "V"), hom("tgt1", "E1", "V"),
    hom("src2", "E2", "V"), hom("tgt2", "E2", "V")
  )
)
codom_sch <- BasicSchema(
  obs = c("V", "E"),
  homs = list(hom("src", "E", "V"), hom("tgt", "E", "V"))
)

cat_dom <- FinCat(schema = dom_sch)
cat_codom <- FinCat(schema = codom_sch)
F <- FinFunctor(
  ob_map = list(V = "V", E1 = "E", E2 = "E"),
  hom_map = list(src1 = "src", tgt1 = "tgt", src2 = "src", tgt2 = "tgt"),
  dom = cat_dom, codom = cat_codom
)

# Source: 3 vertices, 1 E1-edge (1→2), 1 E2-edge (2→3)
source <- ACSet(dom_sch)
add_parts(source, "V", 3)
add_parts(source, "E1", 1)
set_subpart(source, 1, "src1", 1); set_subpart(source, 1, "tgt1", 2)
add_parts(source, "E2", 1)
set_subpart(source, 1, "src2", 2); set_subpart(source, 1, "tgt2", 3)

ResultType <- acset_type(codom_sch)
result <- sigma_migrate(F, source, ResultType)
cat("Combined graph: V =", nparts(result, "V"), ", E =", nparts(result, "E"), "\n")
cat("Edge sources:", subpart(result, NULL, "src"), "\n")
cat("Edge targets:", subpart(result, NULL, "tgt"), "\n")
```

## Summary

| Operation | Function | Description |
|-----------|----------|-------------|
| Terminal | `terminal(schema)` | 1 part per object |
| Initial | `initial(schema)` | 0 parts |
| Product | `product(A, B)` | Cartesian product with projections |
| Coproduct | `coproduct(A, B)` | Disjoint union with injections |
| Pullback | `pullback(f, g)` | Fiber product of cospan |
| Pushout | `pushout(f, g)` | Gluing of span |
| Equalizer | `equalizer(f, g)` | Subobject where f = g |
| Coequalizer | `coequalizer(f, g)` | Quotient identifying f(a) ~ g(a) |
| Homomorphism | `find_homomorphism(P, G)` | Pattern matching |
| Rule | `rule(l, r)` | DPO rewriting rule |
| Rewrite | `rewrite(rule, G)` | Apply rule (auto-match) |
| Delta migrate | `delta_migrate(F, X, T)` | Pullback along functor |
| Sigma migrate | `sigma_migrate(F, X, T)` | Pushforward along functor |
