---
title: "Undirected Wiring Diagrams"
author: "Simon Frost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Undirected Wiring Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**Undirected wiring diagrams** (UWDs) are a categorical formalism for
describing how open systems compose by sharing interfaces. In a UWD, each
subsystem is a *box* with *ports*, and composition happens by wiring
ports of different boxes to common *junctions*. Because the wires are
undirected, junctions model **shared variables** rather than directed
signal flow.

UWDs originate from applied category theory and are a core data structure
in the [AlgebraicJulia](https://www.algebraicjulia.org/) ecosystem. They
are used to compose dynamical systems, agent-based models, Petri nets,
stock-flow models, and more. The key idea is that you can specify the
*architecture* of a composed system (which subsystems share which
variables) separately from the *dynamics* of each subsystem.

For example, an SIR epidemiological model can be decomposed into two
sub-processes—infection and recovery—that share populations through
junctions:

- **Infection** involves the susceptible (`s`) and infected (`i`)
  populations.
- **Recovery** involves the infected (`i`) and recovered (`r`)
  populations.

The junction `i` is shared between both boxes, expressing that the same
infected population participates in both processes.

```{r setup}
library(catlab)
library(acsets)
```

## The UWD schema

A UWD is defined as an ACSet (attributed C-set) over the schema
`SchUWD`. The schema has four object types and three morphisms:

| Object      | Role                                                        |
|-------------|-------------------------------------------------------------|
| `Box`       | A subsystem / process                                       |
| `Port`      | An interface point on a box                                 |
| `OuterPort` | An interface point exposed to the outside world             |
| `Junction`  | A shared variable that ports and outer ports connect to     |

| Morphism          | Signature              | Meaning                              |
|-------------------|------------------------|--------------------------------------|
| `box`             | Port → Box             | Which box a port belongs to          |
| `junction`        | Port → Junction        | Which junction a port is wired to    |
| `outer_junction`  | OuterPort → Junction   | Which junction an outer port exposes |

Junctions also carry a `name` attribute (`name: Junction → Name`) so
that each shared variable has a human-readable label.

```{r schema}
SchUWD
```

An empty UWD is created with `UWD()`:

```{r empty-uwd}
w <- UWD()
nparts(w, "Box")
nparts(w, "Junction")
```

## Creating UWDs with `relation()`

The `relation()` function provides a DSL inspired by relational algebra.
Outer port junction names are passed as bare (unquoted) arguments, and
boxes are specified through a `.boxes` list of `box_spec()` calls.

### Step 1: Define box specifications

Each `box_spec()` takes a name and a set of junction names that the box
connects to:

```{r box-specs}
inf_box <- box_spec("infection", "s", "i")
rec_box <- box_spec("recovery", "i", "r")
inf_box
rec_box
```

### Step 2: Build the UWD with `relation()`

Pass the outer port junction names as bare symbols, and the box
specifications via `.boxes`:

```{r relation-sir}
w_rel <- relation(s, i, r, .boxes = list(
  box_spec("infection", "s", "i"),
  box_spec("recovery", "i", "r")
))
```

This creates a UWD with:

- **3 junctions** named `s`, `i`, `r`
- **3 outer ports** (one for each junction exposed externally)
- **2 boxes** (infection and recovery)
- **4 ports** (infection has 2 ports, recovery has 2 ports)

```{r relation-counts}
nparts(w_rel, "Junction")
nparts(w_rel, "OuterPort")
nparts(w_rel, "Box")
nparts(w_rel, "Port")
```

## Creating UWDs with `uwd()`

The `uwd()` function provides a more concise syntax. The first argument
is a character vector of outer port junction names, and subsequent named
arguments define boxes as character vectors of junction names:

```{r uwd-sir}
w <- uwd(
  c("s", "i", "r"),
  infection = c("s", "i"),
  recovery  = c("i", "r")
)
```

This produces the same structure as the `relation()` call above:

```{r uwd-counts}
nparts(w, "Junction")
nparts(w, "OuterPort")
nparts(w, "Box")
nparts(w, "Port")
```

The `uwd()` syntax is generally preferred for its brevity, while
`relation()` may be more readable when box specifications are complex or
built programmatically.

## Inspecting UWDs

### Counting parts

Use `nparts()` to count elements of each type:

```{r inspect-nparts}
nparts(w, "Box")
nparts(w, "Port")
nparts(w, "Junction")
nparts(w, "OuterPort")
```

### Querying morphisms with `subpart()`

Use `subpart()` to follow morphisms and attributes. For example, to find
which box each port belongs to:

```{r inspect-box}
# Port 1 belongs to which box?
subpart(w, 1, "box")

# Port 3 belongs to which box?
subpart(w, 3, "box")
```

To find which junction each port connects to:

```{r inspect-junction}
# Port 1 connects to which junction?
subpart(w, 1, "junction")

# Port 2 connects to which junction?
subpart(w, 2, "junction")
```

### Reading junction names

Each junction has a `name` attribute:

```{r inspect-names}
# Name of junction 1
subpart(w, 1, "name")

# Name of junction 2
subpart(w, 2, "name")

# Name of junction 3
subpart(w, 3, "name")
```

### Listing all junction names

Use `parts()` to get all part IDs, then retrieve their names:

```{r inspect-all-names}
junc_ids <- parts(w, "Junction")
junction_names <- vapply(junc_ids, function(j) subpart(w, j, "name"), character(1))
junction_names
```

### Querying outer port connections

Each outer port connects to a junction via `outer_junction`:

```{r inspect-outer}
# Outer port 1 connects to which junction?
subpart(w, 1, "outer_junction")

# Outer port 2 connects to which junction?
subpart(w, 2, "outer_junction")
```

## Visualizing UWDs

The `uwd_to_dot()` function generates a Graphviz DOT string for
visualizing a UWD. In the output:

- **Circles** represent junctions (labeled with the junction name)
- **Rectangles** represent boxes
- **Diamonds** represent outer ports (the external interface)
- **Edges** connect boxes to junctions (via ports) and outer ports to
  junctions

```{r viz-sir}
cat(uwd_to_dot(w))
```

If you have the `DiagrammeR` package installed, you can render this
interactively with `to_graphviz(w)`.

## Epidemiological example: SIR model

The classic SIR (Susceptible-Infected-Recovered) compartmental model is
a natural example of system composition. The model consists of two
processes:

1. **Infection**: susceptible individuals become infected through contact
   with infected individuals. This process involves the `s` and `i`
   compartments.
2. **Recovery**: infected individuals recover. This process involves the
   `i` and `r` compartments.

The junction `i` is shared between both processes, expressing that the
infected population is the interface between infection and recovery.

```{r sir}
sir <- uwd(
  c("s", "i", "r"),
  infection = c("s", "i"),
  recovery  = c("i", "r")
)
nparts(sir, "Box")
nparts(sir, "Port")
nparts(sir, "Junction")
```

The DOT visualization shows the two boxes connected through their shared
junction:

```{r sir-dot}
cat(uwd_to_dot(sir))
```

This architecture can be used downstream to compose ODE systems, Petri
nets, or agent-based models for each process into a complete SIR model.

## Predator-prey example: Lotka-Volterra

The Lotka-Volterra predator-prey model decomposes into three processes:

1. **Growth**: prey (`x`) reproduce, depending only on the prey
   population.
2. **Predation**: predators (`y`) consume prey (`x`), involving both
   populations.
3. **Decline**: predators (`y`) die, depending only on the predator
   population.

```{r lotka-volterra}
lv <- uwd(
  c("x", "y"),
  growth    = c("x"),
  predation = c("x", "y"),
  decline   = c("y")
)
```

Let's verify the structure:

```{r lv-inspect}
nparts(lv, "Box")       # 3 boxes: growth, predation, decline
nparts(lv, "Junction")  # 2 junctions: x, y
nparts(lv, "OuterPort") # 2 outer ports
nparts(lv, "Port")      # 4 ports: growth(1) + predation(2) + decline(1)
```

The key feature here is that the `x` junction is shared between growth
and predation, and the `y` junction is shared between predation and
decline. Predation is the process that couples the two populations.

```{r lv-dot}
cat(uwd_to_dot(lv))
```

## Multi-component systems

UWDs naturally express more complex architectures where multiple
subsystems share variables in overlapping ways. Consider an SEIR
(Susceptible-Exposed-Infected-Recovered) epidemiological model with
three processes:

1. **Exposure**: susceptible individuals are exposed through contact
   with infected individuals (involves `s`, `e`, `i`).
2. **Progression**: exposed individuals become infectious (involves `e`,
   `i`).
3. **Recovery**: infected individuals recover (involves `i`, `r`).

```{r seir}
seir <- uwd(
  c("s", "e", "i", "r"),
  exposure    = c("s", "e", "i"),
  progression = c("e", "i"),
  recovery    = c("i", "r")
)
```

```{r seir-inspect}
nparts(seir, "Box")       # 3 boxes
nparts(seir, "Junction")  # 4 junctions: s, e, i, r
nparts(seir, "OuterPort") # 4 outer ports
nparts(seir, "Port")      # 7 ports: exposure(3) + progression(2) + recovery(2)
```

The junction `i` appears in all three processes: as an input to
exposure (infected individuals drive new exposures), as the output of
progression, and as the input to recovery. This is the power of
UWDs—junctions naturally express many-to-many sharing of variables
across subsystems.

```{r seir-dot}
cat(uwd_to_dot(seir))
```

### A larger example: coupled SIR with vital dynamics

Consider two SIR-like processes augmented with vital dynamics (birth and
death). This model has five processes sharing four compartments:

```{r coupled}
coupled <- uwd(
  c("s", "i", "r"),
  birth     = c("s"),
  infection = c("s", "i"),
  recovery  = c("i", "r"),
  waning    = c("r", "s"),
  death     = c("s", "i", "r")
)
```

```{r coupled-inspect}
nparts(coupled, "Box")       # 5 boxes
nparts(coupled, "Junction")  # 3 junctions
nparts(coupled, "OuterPort") # 3 outer ports
nparts(coupled, "Port")      # 9 ports
```

Here the junction `s` participates in four processes (birth, infection,
waning, and death), while `i` participates in three (infection, recovery,
and death). The `death` box connects to all three compartments, showing
how UWDs easily represent processes with broad interfaces.

```{r coupled-dot}
cat(uwd_to_dot(coupled))
```

## Summary

| Function       | Purpose                                      |
|----------------|----------------------------------------------|
| `SchUWD`       | The UWD schema (objects, homs, attrs)        |
| `UWD()`        | Create an empty UWD                          |
| `relation()`   | DSL for building UWDs with `box_spec()`      |
| `box_spec()`   | Define a box with its junction connections   |
| `uwd()`        | Concise syntax for building UWDs             |
| `nparts()`     | Count parts of a given type                  |
| `subpart()`    | Query morphisms and attributes               |
| `parts()`      | List all part IDs of a given type            |
| `uwd_to_dot()` | Generate Graphviz DOT visualization          |

UWDs provide a compositional architecture language: they describe *how*
subsystems connect without prescribing *what* each subsystem does. This
separation of architecture from dynamics is the foundation for
compositional modeling in the AlgebraicJulia ecosystem, and `catlab`
brings this approach to R.
